---
title: "COVID"
author: "Sorabh Kaila"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Import Covid data from Josh Hopkins Github site
```{r}
#install.packages("tidyverse")
library(tidyverse)
```


```{r}
url_in= "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"
```

### Import different datasets...us cases, us deaths, global cases & global deaths.

```{r}
file_names = c("time_series_covid19_confirmed_US.csv",
  "time_series_covid19_confirmed_global.csv",
  "time_series_covid19_deaths_US.csv",
  "time_series_covid19_deaths_global.csv")
  
urls = str_c(url_in,file_names)
urls
```

```{r}
global_cases = read_csv(urls[2])
global_deaths = read_csv(urls[4])
us_cases = read_csv(urls[1])
us_deaths = read_csv(urls[3])
```


### View Global cases & Global deaths data and clean data...delete columns not needed (Lat and Long). Pivot date into 1 column.
```{r}
head(global_cases)
global_cases = subset(global_cases,select = -c(Lat,Long))
global_cases= global_cases%>%
  pivot_longer(cols = -c("Province/State","Country/Region"),
               names_to="date",values_to="cases")
```

```{r}
head(global_deaths)
global_deaths = subset(global_deaths,select = -c(Lat,Long))
global_deaths= global_deaths%>%
  pivot_longer(cols = -c("Province/State","Country/Region"),
               names_to="date",values_to="deaths")
```

### Combine global cases and deaths in 1 dataset and change date format to date
```{r}
library(lubridate)
global = full_join(global_cases,global_deaths,by = c("Province/State","Country/Region","date"))

global = global%>%mutate(global, date = mdy(date))

head(global)
summary(global)
```

### include only data with cases>0
```{r}
global = global%>%filter(cases>0)
summary(global)
```

### Look at outliers
```{r}
global%>%filter(cases>2800000)
```

### US Dataset: Remove columns not needed, pivot date into 1 column and combine cases & deaths into 1 dataset.
```{r}
us_cases = subset(us_cases,select=-c(UID,iso2,iso3,code3,FIPS, Lat, Long_))
us_deaths = subset(us_deaths,select=-c(UID,iso2,iso3,code3,FIPS, Lat, Long_))
us_cases= us_cases%>%
  pivot_longer(cols = -c(Admin2:Combined_Key),
               names_to="date",values_to="cases")
us_deaths= us_deaths%>%
  pivot_longer(cols = -c(Admin2:Population),
               names_to="date",values_to="deaths")

us = full_join(us_cases,us_deaths,by = c("Admin2","Province_State","Country_Region","Combined_Key","date"))

us = us%>%mutate(us, date = mdy(date))
summary(us)
```


### Add combined key and population to Global data so it has the same columns as us data.
```{r}
colnames(global) = c("Province_State","Country_Region","date","cases","deaths")

global = global%>%unite("Combined_Key",c(Province_State,Country_Region),sep=",",na.rm=TRUE,remove=FALSE)
```

### Add global population to global dataset
```{r}
pop_url = "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"

pop = read_csv(pop_url)
pop = subset(pop,select = -c(UID:FIPS))
pop = subset(pop,select = -c(Lat,Long_,Combined_Key))
global = left_join(global,pop,by =c("Province_State","Country_Region"))
```

## Vizualizing Data: US

### Group US data by State. Add death per mil metric.
```{r}
us_state = us %>% group_by(Province_State,Country_Region,date) %>%
summarize(cases=sum(cases),deaths=sum(deaths),Population = sum(Population)) %>% mutate (deaths_per_mil = deaths*1000000/Population) %>% select(Province_State,Country_Region,date,cases,deaths,deaths_per_mil,Population) %>% ungroup()
```

### Group US data by date. Add deaths_per million metric.
```{r}
us_total = us_state %>% group_by(Country_Region,date) %>%
summarize(cases=sum(cases),deaths=sum(deaths),Population = sum(Population)) %>% mutate (deaths_per_mil = deaths*1000000/Population) %>% select(Country_Region,date,cases,deaths,deaths_per_mil,Population) %>% ungroup()
tail(us_total)
```


### Visualize Data: Cases and deaths trend by date
```{r}
us_total %>% 
  filter(cases>0) %>%
  ggplot(aes(x=date, y = cases))+
  geom_line(aes(color="cases"))+
  geom_point(aes(color="cases"))+
  geom_line(aes(y=deaths, color="deaths"))+
  geom_point(aes(y=deaths, color="deaths"))+
  scale_y_log10()
```


### Analyze cases and deaths for NY only
```{r}
us_state %>% 
  filter(Province_State == "New York") %>%
  filter(cases>0) %>%
  ggplot(aes(x=date, y = cases))+
  geom_line(aes(color="cases"))+
  geom_point(aes(color="cases"))+
  geom_line(aes(y=deaths, color="deaths"))+
  geom_point(aes(y=deaths, color="deaths"))+
  scale_y_log10()
```

### Look at the max of us cases and deaths
```{r}
max(us_total$date)
max(us_total$cases)
max(us_total$deaths)
```

### Add...new cases and new deaths
```{r}
us_state = us_state %>% mutate(new_cases = cases-lag(cases))
us_state = us_state %>% mutate(new_deaths = deaths-lag(deaths))
us_total = us_total %>% mutate(new_cases = cases-lag(cases))
us_total = us_total %>% mutate(new_deaths = deaths-lag(deaths))
tail(us_total %>% select(new_cases,new_deaths,everything()))
```

### Plot new cases and new deaths
```{r}
us_state %>%
filter(Province_State == "New York") %>%
filter(new_cases>0) %>%
ggplot(aes(x=date, y = cases))+
geom_line(aes(color="cases"))+
geom_point(aes(color="cases"))+
geom_line(aes(y=deaths, color="deaths"))+
geom_point(aes(y=deaths, color="deaths"))+
scale_y_log10()
```

### Analyze data by State
```{r}
us_state_total = us_state %>% 
  group_by(Province_State) %>%
summarize(cases = max(cases),deaths=max(deaths),Population=max(Population),
          cases_per_thousand = cases*1000/Population,
          deaths_per_thousand = deaths*1000/Population) %>%
  filter(cases>0,Population>0)
```

### States with lowest death rate & max death rate
```{r}
us_state_total %>% slice_min (deaths_per_thousand,n=10) %>% 
  select(deaths_per_thousand,everything())

us_state_total %>% slice_max (deaths_per_thousand,n=10) %>% 
  select(deaths_per_thousand,everything())
```

## Modeling Data...build a linear model to predict deaths per thousand based on cases per thousand.
```{r}
mod = lm(deaths_per_thousand~cases_per_thousand,data = us_state_total)
summary(mod)

us_state_total_pred = us_state_total %>% mutate(pred = predict(mod))

us_state_total_pred %>% ggplot() + geom_point(aes(x = cases_per_thousand,y=deaths_per_thousand),color="blue")+
  geom_point(aes(x = cases_per_thousand,y=pred),color="red")
```
  
## Group global data by country and add new deaths and new cases
```{r}
global_country = global %>% group_by(Country_Region,date) %>%
summarize(cases=sum(cases),deaths=sum(deaths),Population = sum(Population)) %>% mutate (deaths_per_mil = deaths*1000000/Population) %>% select(Country_Region,date,cases,deaths,deaths_per_mil,Population) %>% ungroup()
global_country
global_country = global_country %>% mutate(new_cases = cases-lag(cases))
global_country = global_country %>% mutate(new_deaths = deaths-lag(deaths))
global_country = global_country %>% mutate(new_cases = cases-lag(cases))
global_country = global_country %>% mutate(new_deaths = deaths-lag(deaths))
tail(global_country %>% select(new_cases,new_deaths,everything()))
```

### Create a dataset for totals by Country
```{r}
global_country_total = global_country %>% 
  group_by(Country_Region) %>%
summarize(cases = max(cases),deaths=max(deaths),Population=max(Population),
          cases_per_thousand = cases*1000/Population,
          deaths_per_thousand = deaths*1000/Population) %>%
  filter(cases>0,Population>0)
tail(global_country_total)
```

### Countries with maximum deaths...Peru, Bulgaria and Hungary have the max deaths_per_thousan of population
```{r}
global_country_total %>% slice_max (deaths_per_thousand,n=10) %>% 
  select(deaths_per_thousand,everything())
```

### Countries with minimum deaths...there seems to be reporting issues in the top 3 countries...
### Population is too low for Holy See and Tuvalu so lets discard these. N. Korea shows only 1 case and 6 deaths so defintely there seems to be a reporting issue...Need to understand further
```{r}
global_country_total %>% slice_min (deaths_per_thousand,n=10) %>% 
  select(deaths_per_thousand,everything())
```

### Vizualize data by Country for top 20 countries
```{r}
global_country_top20 = global_country_total %>% slice_max(deaths_per_thousand,n=20) %>% arrange(desc(deaths_per_thousand))

global_country_top20 %>%
ggplot(aes(reorder(x=Country_Region,-deaths_per_thousand),y=deaths_per_thousand))+
  geom_bar(position='dodge',stat = 'identity',color='green',fill='green')+
  ggtitle("Cases by Country")+
  geom_text(aes(label = round(deaths_per_thousand,1)),size=3,color="black",vjust="inward")+
  theme(axis.text.x = element_text(size=6, angle = 90))
```


### Vizualize data by Country for bottom 20 countries
```{r}
global_country_bottom20 = global_country_total %>% slice_min(deaths_per_thousand,n=20) %>% arrange(desc(deaths_per_thousand))

global_country_bottom20 %>%
ggplot(aes(reorder(x=Country_Region,deaths_per_thousand),y=deaths_per_thousand))+
  geom_bar(position='dodge',stat = 'identity',color='green',fill='green')+
  ggtitle("Cases by Country")+
  geom_text(aes(label = round(deaths_per_thousand,2)),size=3,color="black",vjust="inward")+
  theme(axis.text.x = element_text(size=6, angle = 90))
```


### Vizualize data for countries with largest population. This has some interesting observations. US has highest deaths_per thousand and India, Indonesia and Pakisatan are extremely low on the list. Looks like China data is missing from the global dataset.
```{r}
global_country_top_pop = global_country_total %>% slice_max(Population,n=20)

global_country_top_pop %>%
ggplot(aes(reorder(x=Country_Region,-Population),y=deaths_per_thousand))+
  geom_bar(position='dodge',stat = 'identity',color='green',fill='green')+
  ggtitle("Cases by Country")+
  geom_text(aes(label = round(deaths_per_thousand,2)),size=3,color="black",vjust="inward")+
  theme(axis.text.x = element_text(size=6, angle = 90))
```


## Build a linear model for global data
```{r}
mod2 = lm(deaths_per_thousand~cases_per_thousand,data = global_country_total)
summary(mod2)

global_country_total_pred = global_country_total %>% mutate(pred = predict(mod2))

global_country_total_pred %>% ggplot() + geom_point(aes(x = cases_per_thousand,y=deaths_per_thousand),color="blue")+
  geom_point(aes(x = cases_per_thousand,y=pred),color="red")
```

## Bias in the data
### Based on the initial analysis there seems to be a lot of variance between Countries in cases per thousand and deaths per thousand. There may be bias in the measurement system of different countries.
